---
title: "STAT/MATH 495: Problem Set 09"
author: "Luke Haggerty"
date: "2017-11-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)

library(tidyverse)
```



# Collaboration

Please indicate who you collaborated with on this assignment: Individual



# Question 1: Run k-means

#Observations_1

1
```{r}
observations_1 <- read_csv("data/observations_1.csv")

# Set observations to be one of two datasets
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

2
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

3
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

4
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

5
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

6
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```
7
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```
8
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

9
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

10
```{r, echo=FALSE}

observations_1 <- read_csv("data/observations_1.csv")
observations <- observations_1

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

#Observations_2

1
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

2
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

3
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

4
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

5
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

6
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```
7
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

8
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

9
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

10
```{r, echo = FALSE}
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```
**Questions**:

1. Run KMC 10 times on `observations_1` and comment on the consistency of the
results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.
1. Run KMC 10 times on `observations_2` and comment on the consistentcy of the
results.
1. Speculate on the root cause of any consistency or inconsistency in the
results.

**Answers**:

1.The results are not consistent. There are different clusters each time KMC is run. Sometimes the data points are split horizontally, sometimes it is split vertically, sometimes it is split diagonally. 
1. This inconsistency is because the data points are uniformly spread out, so based on the random placement of the first center points, the clusters will group in several different orientations.
1. The results are consistent. Each time the same points are grouped in the same cluster. The only difference is sometimes one cluster is cluster #1, and sometimes it is cluster #2.
1. The results are consistent because the points are already in clear separate groups, and the KMC process will separate these two groups. The reason that sometimes a cluster is labelled #1 and sometimes it is labelled #2 depends on the random placement of the first center points. These center points wil gravitate towards the group they are closest to. This difference does not change the consistency of the results.



# Bonus question: Code your own

Read ISLR page 388 Algorithm 10.1 and implement k-means clustering from scratch.
Don't worry about doing it for general $k$; keep it simple and do it for $k=2$
specifically. Apply it to `observations_2` from above.

